# Maps {#maps}

```{r, echo=FALSE, eval=TRUE, collapse=TRUE, include=FALSE}
library(dplyr)
library(magrittr)
library(LTabundR)
library(swfscDAS)

#data(hiceas2020)
#das <- hiceas2020
#data(settings)
load('data/cruz_proc.RData')
```

To build a flexible system for mapping cruise data, we have the following functions:  

## Base maps {-}  

Begin with a basic map, including EEZ borders:  

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE, fig.width=6, fig.height=4}
m <- map_base(region='cnp')
m
```

We also have a base map for the California Current ... 

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE, fig.width=6, fig.height=4}
m <- map_base(region='ccs')
m
```

And the ETP:  

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE, fig.width=6, fig.height=4}
m <- map_base(region='etp')
m
```

## Add strata {-}  

Add your research strata to your map: 

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE, fig.width=6, fig.height=4}
m <- map_base(region='cnp')
m <- map_strata(m, 
                settings, 
                region='cnp')
m
```

## Add survey tracks {-}  

*Note: this may take a minute to render.*  

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE, fig.width=6, fig.height=4}
m1 <- map_effort(m, cruz)
m1
```

The defaults of `map_effort()` assume, for simplicity, that you want to see the segments to be included in density estimation for the first cohort specified in your settings.  You can adjust this and other defaults using the function arguments. 

## Customizing effort {-}  

### Inputs  {-}  

To demonstrate some of the customization options, consider this map that shows segments to be *excluded* from the `"distance"` (detection function) analysis for our second cohort (`fkw_insular`).

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE, fig.width=6, fig.height=4}
map_effort(m, cruz,
           cohort = 2,
           analysis = 'distance',
           use_type = c(FALSE),
           effort_color='firebrick',
           effort_stroke=2.5,
           effort_linetype=1,)
```


### Color-code conditions {-}  

Your second customization option is to add format variables to the `segments` slot of the cohort of interest in the `cruz` object. This gives you full control of line color, thickness, and line-type according to whatever specifications you wish to set, e.g., color-coding by effort type or Beaufort sea state.  

This is possible because the function `map_effort()` looks for the variables `col` (line color), `lwd` (line thickness or stroke), and `lty` (line type) in the columns of `cruz$segments`. If these columns exist, the values therein will be used instead of the function defaults.  

For example, color-code by Beaufort scale:

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE, fig.width=6, fig.height=4}
# Save copy of cruz object data to modify
cruz2 <- cruz
segments <- cruz2$cohorts$default$density$segments

# Add column `col`: color code by BFT sea state
bft_colors <- c('steelblue4','steelblue2','cadetblue1','grey')
segments$col <- bft_colors[4]
segments$col[ segments$avgBft <= 7 ] <- bft_colors[3] # bft 5 +
segments$col[ segments$avgBft <= 4 ] <- bft_colors[2] # bft 3 - 4
segments$col[ segments$avgBft <= 2 ] <- bft_colors[1] # bft 0 -2

# Update sub_segments slot in `cruz` object
cruz2$cohorts$default$density$segments <- segments

# Update map 
m_custom2 <- map_effort(m, cruz2)

# Add legend using native functions from mapping package `tmap`
m_custom2 <- 
  m_custom2 + 
  tmap::tm_add_legend('line', 
	                    col = bft_colors,
	                    lwd = 3,
	                    labels = c(' 0 - 2', 
	                               ' 3 - 4', 
	                               ' 5 +', 
	                               ' no data'),
	                     title="Beaufort sea state") +
  tmap::tm_layout(legend.position=c('left','bottom'))

# Show map
m_custom2
```

