# (PART) Case studies {-}

# WHICEAS {#whiceas}

Here we demonstrate code that reproduces the Bradford et al. (2022) WHICEAS report within the new `LTabundR` framework. This study estimates cetacean abundance for Hawaiian WHICEAS study area for 2017 and 2020. Here we use survey data from 1986 to 2020 to estimate *Relative g(0)* and detection functions. Currently, coefficients of variation (CV) of density and abundance are estimated using only 100 bootstrap iterations (the publication uses 1,000) to reduce processing time.  

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, include=TRUE, warning=FALSE, message=FALSE}
library(dplyr)
library(LTabundR)
```

```{r, echo=FALSE, eval=TRUE, collapse=TRUE, include=TRUE, warning=FALSE, message=FALSE}
reproc_from_scratch <- FALSE
rerun_from_scratch <- FALSE
g0_scenario <- 3
# If you want to confirm the package works, change to TRUE.
# If you already know it does, change to FALSE so that knitting is quick
# (it will use saved R objects instead of producing them anew)
```

## Data processing {-}  

### Settings {-}  

#### Survey-wide settings {-}

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE}
data(species_codes)
data(ships)
data(group_size_coefficients)
edits <- readRDS('cnp_1986_2020_edits.RData')

survey <- load_survey_settings(
  out_handling = 'remove',
  max_row_interval = Inf,
  segment_method = "equallength",
  segment_target_km = 150,
  segment_max_interval = 24,
  segment_remainder_handling = c("segment"),
  ship_list = ships,
  species_codes = species_codes,
  group_size_coefficients = group_size_coefficients,
  smear_angles = FALSE
)
```

#### Geostrata {-}  

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE}
data(strata_cnp)
```

#### Cohort-specific settings {-} 

##### Cohort 1: all species {-} 

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE}
all_species <- load_cohort_settings(
  id = "all", # *
  species = NULL, 
  strata = c('WHICEAS', 'HI_EEZ', 'OtherCNP'), # *
  probable_species = FALSE,
  sighting_method = 0,
  cue_range = 0:7,
  school_size_range = c(0, 10000),
  school_size_calibrate = TRUE,
  calibration_floor = 0,
  use_low_if_na = TRUE,
  io_sightings = 0,
  geometric_mean_group = TRUE,
  truncation_km = 7.5, # *
  beaufort_range = 0:6,
  abeam_sightings = TRUE,
  strata_overlap_handling = c("smallest"),
  distance_types = c('S','F','N'),
  distance_modes = c('P','C'),
  distance_on_off = TRUE
)
```

##### Cohort 2: bottlenose dolphins {-} 

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE}
bottlenose <- load_cohort_settings(
  id = "bottlenose",
  species = c('015', '018', '021', '032'),
  strata = c('WHICEAS', 'HI_EEZ', 'OtherCNP',
             'Bottlenose_BI', 'Bottlenose_OUFI', 'Bottlenose_KaNi'),
  truncation_km = 7.5)
```

##### Cohort 3: pantropical spotted dolphins {-} 

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE}
spotted <- load_cohort_settings(
  id = "spotted",
  species = '002',
  strata = c('WHICEAS', 'HI_EEZ', 'OtherCNP',
             'Spotted_OU','Spotted_FI','Spotted_BI'),
  truncation_km = 7.5)
```

### Process {-}  

```{r, echo=TRUE, eval=reproc_from_scratch, collapse=TRUE, message=FALSE, warning=FALSE}
settings <- load_settings(strata = strata_cnp,
                          survey = survey,
                          cohorts = list(all_species,
                                         bottlenose,
                                         spotted))

cruz <- 
  process_surveys(das_file = 'data/surveys/CenPac1986-2020_Final_alb.das',
                  settings = settings,
                  edits = edits) 

save(cruz, file='whiceas/whiceas_cruz.RData')
```

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE}
load('whiceas/whiceas_cruz.RData')
```

## Data exploration {-}  

### Processed data structure {-}  

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE}
cruz_structure(cruz)
```

### Map of sightings {-}  

For the entire `cruz` object:  

```{r, echo=TRUE, eval=FALSE, collapse=TRUE, message=FALSE, warning=FALSE, fig.height=7, fig.width=7}
map_cruz(cruz, sightings_color = 'firebrick')
```

For just the surveys of interest:  

```{r, echo=TRUE, eval=FALSE, collapse=TRUE, message=FALSE, warning=FALSE, fig.height=7, fig.width=7}
# Filter
cruz1720 <- filter_cruz(cruz, 
                        years = c(2017, 2020), 
                        regions = 'WHICEAS',
                        eff_types = 'S',
                        bft_range = 0:6)

# Map, including survey tracks
map_cruz(cruz1720,
         sightings_color = 'firebrick', 
         effort_show = TRUE, 
         effort_resolution = 3,
         effort_weight = 4,
         effort_opacity = .4)
```

### Interactive dashboard {-}  

Use this to determine truncation distances.  Note them in your `lta()` code below: 

```{r, echo=TRUE, eval=FALSE, collapse=TRUE, message=FALSE, warning=FALSE}
cruz_explorer(cruz)
```

## Rg0 {-}  

We can use the built-in dataset, `data(g0_results)`, which has Beaufort-specific *Relative g(0)* estimates for most species based on 1986-2020 surveys.  

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE}
data("g0_results")
Rg0 <- g0_results
```

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE, fig.width=7, fig.height=9}
# Plot the results: 
g0_plot(Rg0, panes = 3)
```

To explore the effects of `LTabundR`'s *g(0)* estimation routines on our abundance estimates, we will be running the analyses below with three different *g(0)* scenarios. In scenario 1, we will manually specify the weighted `g(0)` and its CV for each species in 2020; these results should be nearly exact replicates of those from Bradford et al. (2021), and we can be sure that any discrepancy between the two sets of results are not attributable the *g(0)* aspect of the analysis. Second, we will allow `LTabundR` to calculate the weighted `g(0)` and its CV using the same *Relative g(0)* values that were used in Bradford et al. (2021) (i.e., the results of Barlow (2015), which used surveys from 1986 to 2010 and is provided in `LTabundR` as a built-in dataset), such that any difference between scenarios 1 and 2 are likely due to the weighted *g(0)* subroutines. Finally, in scenario 3, we will use the *Relative g(0)* esitmates produced by `LTabundR` using a more extensive survey dataset (1986-2020). 

For each *g(0)* scenario, our `Rg0` object and `results_path` will differ:

```{r, echo=TRUE, eval=FALSE, collapse=TRUE, message=FALSE, warning=FALSE}
# Specify scenario number here
g0_scenario <- 1
```

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE}
# Set Rg0 source and results path
if(g0_scenario == 1){ # Using manually-specified g0 and its CV
  Rg0 <- NULL
  results_path <- 'whiceas/lta_manual/'
}
if(g0_scenario == 2){ # Using Rg0 estimates from Barlow 2015
  data(barlow_2015)
  Rg0 <- barlow_2015
  results_path <- 'whiceas/lta_barlow/'
}
if(g0_scenario == 3){ # New LTabundR estimates of Rg0
  data(g0_results)
  Rg0 <- g0_results
  results_path <- 'whiceas/lta/'
}
```

## Density & abundance {-}  

First we can define common values that will be constant across all estimates we produce: 

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE}
bootstraps <- 200
years <- 1986:2020
fit_regions <- NULL
fit_not_regions <- NULL
toplot = TRUE
verbose = TRUE

df_settings <-
  list(covariates = c('bft','lnsstot','cruise','year','ship','species'),
       covariates_factor = c(FALSE, FALSE, TRUE, TRUE, TRUE, TRUE),
       covariates_levels = 2,
       covariates_n_per_level = 10,
       simplify_cue = TRUE,
       simplify_bino = TRUE,
       detection_function_base = 'hn',
       base_model = '~1',
       delta_aic = 2)
```

For most species, we want to estimate density/abundance for the same set of year-region scenarios. To reduce code redundancy, as well as the risk of typing errors (and our work!), we can use the `LTabundR` function `lta_estimates()` to economize how we prepare our `estimates` input.  

For most species, these are the year-region scenarios for which we want estimates:  

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE}
scenarios <- list(list(years = 2017,
                       regions = 'WHICEAS'),
                  list(years = 2020, 
                       regions = 'WHICEAS'))
```

The `lta_estimates()` function will generate a custom function that makes it easy to create a set of `estimates` sub-lists for each species of interest:  

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE}
estimator <- lta_estimates(scenarios)
```

That result, `estimator`, is actually a function. Here's an example of how this function will work, using the first species pool as an example: 

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE}
estimates <-
  c(estimator(spp = '013', title = "Striped dolphin"),
    estimator(spp = '026', title = "Fraser's dolphin", alt_g0_spp = '013'),
    estimator(spp = '031', title = "Melon-headed whale", alt_g0_spp = '013'))

estimates  
```

The output of `estimator()` is a list of sub-lists specifying a set of density/abundance estimates you want to produce based on the detection function for a single species pool.  

Here is the full code for producing those estimates for all species from Bradford et al. (2021):  

### Multi-species pool 1 {-} 

```{r, echo=TRUE, eval=rerun_from_scratch, collapse=TRUE, message=FALSE, warning=FALSE}
# Striped dolphin (013), Fraser's dolphin (026), Melon-headed whale (031)

if(TRUE){ # toggle

  # Detection function specifications
  fit_filters <-
    list(spp = c('013', '026', '031'),
         pool = 'Multi-species pool 1',
         cohort = 'all',
         truncation_distance = 5,
         other_species = 'remove',
         years = years,
         regions = fit_regions,
         not_regions = fit_not_regions)

  # Density / abundance estimation plan
  estimates <-
    c(estimator(spp = '013', title = "Striped dolphin"),
      estimator(spp = '026', title = "Fraser's dolphin", alt_g0_spp = '013'),
      estimator(spp = '031', title = "Melon-headed whale", alt_g0_spp = '013'))
  
  estimates
  
  # Manually specify g0 and its CV -- only for g0_scenario 1
  # Two specifications per species, one for 2017 and one for 2020
  if(g0_scenario==1){
    # Striped 
    estimates[[1]]$g0 <- 0.35; estimates[[1]]$g0_cv <- 0.19
    estimates[[2]]$g0 <- 0.31; estimates[[2]]$g0_cv <- 0.22
    # Fraser's
    estimates[[3]]$g0 <- 0.35; estimates[[3]]$g0_cv <- 0.10
    estimates[[4]]$g0 <- 0.31; estimates[[4]]$g0_cv <- 0.22
    # Melon-headed
    estimates[[5]]$g0 <- 0.35; estimates[[5]]$g0_cv <- 0.19
    estimates[[6]]$g0 <- 0.31; estimates[[6]]$g0_cv <- 0.22
  }
  
  # Run analysis
  results <- lta(cruz, Rg0, fit_filters, df_settings, estimates,
                 bootstraps = bootstraps,
                 toplot=toplot, verbose=verbose)

  # Save result
  (results_file <- paste0(results_path, fit_filters$pool, '.RData'))
  saveRDS(results, file=results_file)
}
```

### Multi-species pool 2 {-} 

```{r, echo=TRUE, eval=rerun_from_scratch, collapse=TRUE, message=FALSE, warning=FALSE}
# Rough-toothed dolphin (15), Common bottlenose dolphin (18), Risso's (21),
# Pygmy killer whale (32)

# Notes
# Bottlenose abundance is estimated in a separate cohort, but included here for DF fitting

if(TRUE){ # toggle

  # Detection function specifications
  fit_filters <-
    list(spp = c('015', '018', '021', '032'),
         pool = 'Multi-species pool 2',
         cohort = 'all',
         truncation_distance = 5,
         years = years,
         regions = fit_regions,
         not_regions = fit_not_regions)

  # Density / abundance estimation plan
  estimates <-
      c(estimator(spp = '015', title = "Rough-toothed dolphin"),
        estimator(spp = '021', title = "Risso's dolphin"),
        estimator(spp = '032', title = "Pygmy killer whale"))

  if(g0_scenario==1){
    # Rough-toothed 
    estimates[[1]]$g0 <- 0.09; estimates[[1]]$g0_cv <- 0.45
    estimates[[2]]$g0 <- 0.07; estimates[[2]]$g0_cv <- 0.51
    # Risso's
    estimates[[3]]$g0 <- 0.57; estimates[[3]]$g0_cv <- 0.18
    estimates[[4]]$g0 <- 0.52; estimates[[4]]$g0_cv <- 021
    # Pygmy killer
    estimates[[5]]$g0 <- 0.14; estimates[[5]]$g0_cv <- 0.25
    estimates[[6]]$g0 <- 0.11; estimates[[6]]$g0_cv <- 0.28
  }
  
  # Run analysis
  results <- lta(cruz, Rg0, fit_filters, df_settings, estimates,
                 use_g0 = TRUE,
                 bootstraps = bootstraps,
                 toplot=toplot, verbose=verbose)

  # Save result
  (results_file <- paste0(results_path, fit_filters$pool, '.RData'))
  saveRDS(results, file=results_file)
}
```

### Multi-species pool 3 {-} 

```{r, echo=TRUE, eval=rerun_from_scratch, collapse=TRUE, message=FALSE, warning=FALSE}
# Short-finned pilot whale (036), Longman's beaked whale (065)

# No Rg(0) available for Longman's -- will use SF pilot whale instead to estimate its weighted g0

if(TRUE){ # toggle

  # Detection function specifications
  fit_filters <-
    list(spp = c('036', '065'),
         pool = 'Multi-species pool 3',
         cohort = 'all',
         truncation_distance = 5,
         years = years,
         regions = fit_regions,
         not_regions = fit_not_regions)

  # Density / abundance estimation plan
  estimates <-
    c(estimator(spp = '036', title = "Short-finned pilot whale"),
      estimator(spp = '065', title = "Longman's beaked whale", 
                alt_g0_spp = '036'))

  if(g0_scenario==1){
    # Short-finned 
    estimates[[1]]$g0 <- 0.58; estimates[[1]]$g0_cv <- 0.15
    estimates[[2]]$g0 <- 0.52; estimates[[2]]$g0_cv <- 0.19
    # Longman's
    estimates[[3]]$g0 <- 0.58; estimates[[3]]$g0_cv <- 0.15
    estimates[[4]]$g0 <- 0.52; estimates[[4]]$g0_cv <- 0.19
  }
  
  # Run analysis
  results <- lta(cruz, Rg0, fit_filters, df_settings, estimates,
                 use_g0 = TRUE,
                 bootstraps = bootstraps,
                 toplot=toplot, verbose=verbose)

  # Save result
  (results_file <- paste0(results_path, fit_filters$pool, '.RData'))
  saveRDS(results, file=results_file)
}
```

### Multi-species pool 4 {-} 

```{r, echo=TRUE, eval=rerun_from_scratch, collapse=TRUE, message=FALSE, warning=FALSE}
# Killer whale (37), sperm whale (46)

if(TRUE){ # toggle

  # Detection function specifications
  fit_filters <-
    list(spp = c('037', '046'),
         pool = 'Multi-species pool 4',
         cohort = 'all',
         truncation_distance = 5.5,
         other_species = 'remove',
         years = years,
         regions = fit_regions,
         not_regions = fit_not_regions)

  # Density / abundance estimation plan
  estimates <-
      c(estimator(spp = '037', title = "Killer whale"),
        estimator(spp = '046', title = "Sperm whale"))

  if(g0_scenario==1){
    # Killer (no sightings in ALB et al 2021)
    estimates[[1]]$g0 <- 1; estimates[[1]]$g0_cv <- 0
    estimates[[2]]$g0 <- 1; estimates[[2]]$g0_cv <- 0
    # Sperm
    estimates[[3]]$g0 <- 0.63; estimates[[3]]$g0_cv <- 0.34
    estimates[[4]]$g0 <- 0.61; estimates[[4]]$g0_cv <- 0.37
  }
  
  # Run analysis
  results <- lta(cruz, Rg0, fit_filters, df_settings, estimates,
                 use_g0 = TRUE,
                 bootstraps = bootstraps,
                 toplot=toplot, verbose=verbose)

  # Save result
  (results_file <- paste0(results_path, fit_filters$pool, '.RData'))
  saveRDS(results, file=results_file)
}
```

### Multi-species pool 5 {-} 

```{r, echo=TRUE, eval=rerun_from_scratch, collapse=TRUE, message=FALSE, warning=FALSE}
# Pygmy sperm whale (47), dwarf sperm whale (48), UNID Kogia (80),
# Blainville's beaked whale (59), Cuvier's beaked whale (61),
# UNID Mesoplodon (51), UNID beaked whale (49), Minke whale (71)

if(TRUE){ # toggle

  # Detection function specifications
  fit_filters <-
    list(spp = c('047', '048', '080', '059', '061', '051', '049', '071'),
         pool = 'Multi-species pool 5',
         cohort = 'all',
         truncation_distance = 4.5,
         years = years,
         regions = fit_regions,
         not_regions = fit_not_regions)

  # Density / abundance estimation plan
  estimates <-
      c(estimator(spp = '047', title = "Pygmy sperm whale"),
        estimator(spp = '048', title = "Dwarf sperm whale"),
        estimator(spp = '080', title = "Unidentified Kogia"),
        estimator(spp = '059', title = "Blainville's beaked whale"),
        estimator(spp = '061', title = "Cuvier's beaked whale"),
        estimator(spp = '051', title = "Unidentified Mesoplodon"),
        estimator(spp = '049', 
                  title = "Unidentified beaked whale",
                  alt_g0_spp = c('061','051'),
                  combine_g0 = TRUE),
        estimator(spp = '071', title = "Minke whale"))

  # Note Barlow2015 provides absolute estimates for Cuviers and UNID Mesop
  
  if(g0_scenario==1){
    # Pygmy sperm (no sightings in ALB, using other kogia values)
    estimates[[1]]$g0 <- 0.005; estimates[[1]]$g0_cv <- 0.15
    estimates[[2]]$g0 <- 0.004; estimates[[2]]$g0_cv <- 0.15
    # Dwarf sperm (no ALB sightings in 2017, using 2020 g0)
    estimates[[3]]$g0 <- 0.005; estimates[[3]]$g0_cv <- 0.15
    estimates[[4]]$g0 <- 0.004; estimates[[4]]$g0_cv <- 0.15
    # UNID Kogia (no ALB sightings 2020, using 2017 g0)
    estimates[[5]]$g0 <- 0.005; estimates[[5]]$g0_cv <- 0.15
    estimates[[6]]$g0 <- 0.004; estimates[[6]]$g0_cv <- 0.15
    # Blainville's (no ALB sightings 2017, using 2020 g0)
    estimates[[7]]$g0 <- 0.11; estimates[[7]]$g0_cv <- 0.30
    estimates[[8]]$g0 <- 0.11; estimates[[6]]$g0_cv <- 0.30
    # Cuvier's (no sightings in ALB -- using unid beaked g0)
    estimates[[9]]$g0 <- 0.13; estimates[[9]]$g0_cv <- 0.20
    estimates[[10]]$g0 <- 0.11; estimates[[10]]$g0_cv <- 0.21
    # UNID Mesop (no ALB sightings 2017, using 2020 g0)
    estimates[[11]]$g0 <- 0.11; estimates[[11]]$g0_cv <- 0.30
    estimates[[12]]$g0 <- 0.11; estimates[[12]]$g0_cv <- 0.30
    # UNID beaked
    estimates[[13]]$g0 <- 0.13; estimates[[13]]$g0_cv <- 0.20
    estimates[[14]]$g0 <- 0.11; estimates[[14]]$g0_cv <- 0.21
  }
  
  # Run analysis
  results <- lta(cruz, Rg0, fit_filters, df_settings, estimates,
                 use_g0 = TRUE,
                 bootstraps = bootstraps,
                 toplot=toplot, verbose=verbose)

  # Save result
  (results_file <- paste0(results_path, fit_filters$pool, '.RData'))
  saveRDS(results, file=results_file)
}
```

### Multi-species pool 6 {-} 

```{r, echo=TRUE, eval=rerun_from_scratch, collapse=TRUE, message=FALSE, warning=FALSE}
# Bryde's whale (72), Sei whale (73), Fin whale (74), Blue whale (75),
# Sei/Bryde's (99), Fin/Sei/Bryde's (72, 73, 74, 99)

# Bryde's, Sei's, and Sei/Bryde's all use same Rg0 (title = "Sei/Bryde's")
# Sei/Bryde's/Fin use an average of Fin and Sei/Bryde's.

if(TRUE){ # toggle

  # Detection function specifications
  fit_filters <-
    list(spp = c('072', '073', '074','075','099'),
         pool = 'Multi-species pool 6',
         cohort = 'all',
         truncation_distance = 5.0,
         years = years,
         regions = fit_regions,
         not_regions = fit_not_regions)

  # Density / abundance estimation plan
   estimates <-
      c(estimator(spp = '072', title = "Bryde's whale"),
        estimator(spp = '073', title = "Sei whale"),
        estimator(spp = '074', title = "Fin whale"),
        estimator(spp = '075', title = "Blue whale"),
        estimator(spp = '099', title = "Sei/Bryde's whale"),
        estimator(spp = c('072', '073', '099', '074'), 
                  title = "Sei/Bryde's/Fin whale",
                  combine_g0 = TRUE))
  
   if(g0_scenario==1){
    # Brydes (no ALB sightings -- using sei values)
    estimates[[1]]$g0 <- 0.38; estimates[[1]]$g0_cv <- 0.21
    estimates[[2]]$g0 <- 0.38; estimates[[2]]$g0_cv <- 0.21
    # Sei (no ALB sightings in 2017, using 2020 g0)
    estimates[[3]]$g0 <- 0.38; estimates[[3]]$g0_cv <- 0.21
    estimates[[4]]$g0 <- 0.38; estimates[[4]]$g0_cv <- 0.21
    # Fin (no ALB sightings in 2017, using 2020 g0)
    estimates[[5]]$g0 <- 0.30; estimates[[5]]$g0_cv <- 0.29
    estimates[[6]]$g0 <- 0.30; estimates[[6]]$g0_cv <- 0.29
    # Blue (no ALB sightings, using fin values)
    estimates[[7]]$g0 <- 0.30; estimates[[7]]$g0_cv <- 0.29
    estimates[[8]]$g0 <- 0.30; estimates[[8]]$g0_cv <- 0.29
    # Sei / Bryde's (no ALB sightings in 2017, using 2020 g0)
    estimates[[9]]$g0 <- 0.38; estimates[[9]]$g0_cv <- 0.21
    estimates[[10]]$g0 <- 0.38; estimates[[10]]$g0_cv <- 0.21
    # Sei / Bryde's / Fin (no ALB sightings in 2017, using 2020 g0)
    estimates[[11]]$g0 <- 0.34; estimates[[11]]$g0_cv <- 0.17
    estimates[[12]]$g0 <- 0.34; estimates[[12]]$g0_cv <- 0.17
  }
  
  # Run analysis
  results <- lta(cruz, Rg0, fit_filters, df_settings, estimates,
                 use_g0 = TRUE,
                 bootstraps = bootstraps,
                 toplot=toplot, verbose=verbose)

  # Save result
  (results_file <- paste0(results_path, fit_filters$pool, '.RData'))
  saveRDS(results, file=results_file)
}
```

### Humpback whale {-} 

```{r, echo=TRUE, eval=rerun_from_scratch, collapse=TRUE, message=FALSE, warning=FALSE}
if(TRUE){ # toggle

  # Detection function specifications
  fit_filters <-
    list(spp = c('076'),
         pool = 'Humpback whale',
         cohort = 'all',
         truncation_distance = 5.5,
         years = years,
         regions = fit_regions,
         not_regions = fit_not_regions)

  # Density / abundance estimation plan
   estimates <-c(estimator(spp = '076', title = "Humpback whale"))
   
   if(g0_scenario==1){
    # No ALB sightings in 2017, using 2020 g0
    estimates[[1]]$g0 <- 0.68; estimates[[1]]$g0_cv <- 0.36
    estimates[[2]]$g0 <- 0.68; estimates[[2]]$g0_cv <- 0.36
   }
  
  # Run analysis
  results <- lta(cruz, Rg0, fit_filters, df_settings, estimates,
                 use_g0 = TRUE,
                 bootstraps = bootstraps,
                 toplot=toplot, verbose=verbose)

  # Save result
  (results_file <- paste0(results_path, fit_filters$pool, '.RData'))
  saveRDS(results, file=results_file)
}
```

### Unidentified rorquals {-} 

```{r, echo=TRUE, eval=rerun_from_scratch, collapse=TRUE, message=FALSE, warning=FALSE}
# UNID rorquals (70)

if(TRUE){ # toggle

  # Detection function specifications
  fit_filters <-
    list(spp = c('070'),
         pool = 'Unidentified rorqual',
         cohort = 'all',
         truncation_distance = 5.5,
         years = years,
         regions = fit_regions,
         not_regions = fit_not_regions)

  # Density / abundance estimation plan
  estimates <-
      c(estimator(spp = '070', 
                  title = "Unidentified rorqual",
                  alt_g0_spp = c('071','099','074','075'),
                  combine_g0 = TRUE))
  
  if(g0_scenario==1){
    estimates[[1]]$g0 <- 0.35; estimates[[1]]$g0_cv <- 0.18
    estimates[[2]]$g0 <- 0.32; estimates[[2]]$g0_cv <- 0.20
  }
  
  # Run analysis
  results <- lta(cruz, Rg0, fit_filters, df_settings, estimates,
                 use_g0 = TRUE,
                 bootstraps = bootstraps,
                 toplot=toplot, verbose=verbose)

  # Save result
  (results_file <- paste0(results_path, fit_filters$pool, '.RData'))
  saveRDS(results, file=results_file)
}
```

### Unidentified dolphins {-} 

```{r, echo=TRUE, eval=rerun_from_scratch, collapse=TRUE, message=FALSE, warning=FALSE}
# UNID dolphin (177, 277, 377, 77)

if(TRUE){ # toggle

  spp <- c('177','277','377','077')
  pool_title <- 'Unidentified dolphin'

  # Detection function specifications
  fit_filters <-
    list(spp = c('177','277','377','077'),
         pool = pool_title,
         cohort = 'all',
         truncation_distance = 5.5,
         other_species = 'coerce',
         years = years,
         regions = fit_regions,
         not_regions = fit_not_regions)

  # Density / abundance estimation plan
  estimates <- estimator(spp = spp, 
                         title = pool_title,
                         alt_g0_spp = c('002','013','018','015', '036', '021'),
                         combine_g0 = TRUE)
  
  if(g0_scenario==1){
    estimates[[1]]$g0 <- 0.33; estimates[[1]]$g0_cv <- 0.08
    estimates[[2]]$g0 <- 0.29; estimates[[2]]$g0_cv <- 0.10
  }
  
  # Run analysis
  results <- lta(cruz, Rg0, fit_filters, df_settings, estimates,
                 use_g0 = TRUE,
                 bootstraps = bootstraps,
                 toplot=toplot, verbose=verbose)

  # Save result
  (results_file <- paste0(results_path, fit_filters$pool, '.RData'))
  saveRDS(results, file=results_file)
}
```

### Unidentified cetaceans {-} 

```{r, echo=TRUE, eval=rerun_from_scratch, collapse=TRUE, message=FALSE, warning=FALSE}
# UNID cetacean (78, 79, 98, 96)

if(TRUE){ # toggle

  spp <- c('078','079','098','096')
  pool_title <- 'Unidentified cetacean'

  # Detection function specifications
  fit_filters <-
    list(spp = spp,
         pool = pool_title,
         cohort = 'all',
         truncation_distance = 5.5,
         other_species = 'coerce',
         years = years,
         regions = fit_regions,
         not_regions = fit_not_regions)

  # Density / abundance estimation plan
  estimates <- estimator(spp = spp, 
                         title = pool_title, 
                         g0=1.0, 
                         g0_cv = 0.0)

  if(g0_scenario==1){
    estimates[[1]]$g0 <- 1.0; estimates[[1]]$g0_cv <- 0.0
    estimates[[2]]$g0 <- 1.0; estimates[[2]]$g0_cv <- 0.0
  }
  
  # Run analysis
  results <- lta(cruz, Rg0, fit_filters, df_settings, estimates,
                 use_g0 = TRUE,
                 bootstraps = bootstraps,
                 toplot=toplot, verbose=verbose)

  # Save result
  (results_file <- paste0(results_path, fit_filters$pool, '.RData'))
  saveRDS(results, file=results_file)
}
```

### Bottlenose dolphin {-} 

```{r, echo=TRUE, eval=rerun_from_scratch, collapse=TRUE, message=FALSE, warning=FALSE}
# Bottlenose dolphin (018)

if(TRUE){ # toggle

  # Detection function specifications
  fit_filters <-
    list(spp = c('015', '018', '021', '032'),
         pool = 'Bottlenose dolphin',
         cohort = 'bottlenose',
         truncation_distance = 5,
         years = years,
         regions = fit_regions,
         not_regions = fit_not_regions)

  # Density / abundance estimation plan
  scenarios <- list(
    list(years = 2017,
       regions = 'WHICEAS',
       regions_remove = c('Bottlenose_KaNi', 'Bottlenose_OUFI', 'Bottlenose_BI'),
       region_title = '(WHICEAS)'),
    list(years = 2020,
       regions = 'WHICEAS',
       regions_remove = c('Bottlenose_KaNi', 'Bottlenose_OUFI', 'Bottlenose_BI'),
       region_title = '(WHICEAS)'))
  
  estimator <- lta_estimates(scenarios)
  estimates <- estimator(spp = '018', title = 'Bottlenose dolphin')
  
  if(g0_scenario==1){
    # No ALB sightings in 2017, using 2020 g0
    estimates[[1]]$g0 <- 0.24; estimates[[1]]$g0_cv <- 0.38
    estimates[[2]]$g0 <- 0.24; estimates[[2]]$g0_cv <- 0.38
  }
  
  # Run analysis
  results <- lta(cruz, Rg0, fit_filters, df_settings, estimates,
                 use_g0 = TRUE,
                 bootstraps = bootstraps,
                 toplot=toplot, verbose=verbose)

  # Save result
  (results_file <- paste0(results_path, fit_filters$pool, '.RData'))
  saveRDS(results, file=results_file)
}
```

### Pantropical spotted dolphin {-} 

```{r, echo=TRUE, eval=rerun_from_scratch, collapse=TRUE, message=FALSE, warning=FALSE}
# Pantropical spotted dolphin (002)

if(TRUE){ # toggle

  # Detection function specifications
  fit_filters <-
    list(spp = c('002'),
         pool = 'Pantropical spotted dolphin',
         cohort = 'spotted',
         truncation_distance = 5,
         years = years,
         regions = fit_regions,
         not_regions = fit_not_regions)

  # Density / abundance estimation plan
  scenarios <- list(
    list(years = 2017,
       regions = 'WHICEAS',
       regions_remove = c('Spotted_OU', 'Spotted_FI', 'Spotted_BI'),
       region_title = '(WHICEAS)'),
    list(years = 2020,
       regions = 'WHICEAS',
       regions_remove = c('Spotted_OU', 'Spotted_FI', 'Spotted_BI'),
       region_title = '(WHICEAS)'))
  
  estimator <- lta_estimates(scenarios)
  estimates <- estimator(spp = '002', title = 'Pantropical spotted dolphin')
  
  if(g0_scenario==1){
    estimates[[1]]$g0 <- 0.28; estimates[[1]]$g0_cv <- 0.11
    estimates[[2]]$g0 <- 0.25; estimates[[2]]$g0_cv <- 0.13
  }
  
  # Run analysis
  results <- lta(cruz, Rg0, fit_filters, df_settings, estimates,
                 use_g0 = TRUE,
                 bootstraps = bootstraps,
                 toplot=toplot, verbose=verbose)

  # Save result
  (results_file <- paste0(results_path, fit_filters$pool, '.RData'))
  saveRDS(results, file=results_file)
}
```

## Results {-}

To review results, we will use *g(0)* scenario 3 (new auto-generated estimates of *Relative g(0)* from `LTabundR`). 

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE}
# Load results
ltas <- lta_enlist('whiceas/lta/')
```

### Tables {-}  

Generate report: 

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE}
reporti <- lta_report(ltas, cruz)
```

**Table 1a.** *Total and systematic sample sizes.* 

```{r, echo=TRUE, eval=FALSE, collapse=TRUE, message=FALSE, warning=FALSE}
reporti$table1a 
```

```{r, echo=FALSE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE}
reporti$table1a %>% 
DT::datatable(extensions = "FixedColumns",
              options=list(scrollX=TRUE,
                           scrollY='450px',
                           paging = FALSE,
                           fixedColumns = list(leftColumns = 2),
                           initComplete = htmlwidgets::JS(
    "function(settings, json) {$(this.api().table().container()).css({'font-size': '9pt'});}")
  ))
```

--- 

**Table 1b.** *Sample sizes of sightings used in density esitmation.*  

```{r, echo=TRUE, eval=FALSE, collapse=TRUE, message=FALSE, warning=FALSE}
reporti$table1b 
```

```{r, echo=FALSE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE}
reporti$table1b %>% 
DT::datatable(extensions = "FixedColumns",
              options=list(scrollX=TRUE,
                           scrollY='450px',
                           paging = FALSE,
                           fixedColumns = list(leftColumns = 2),
                           initComplete = htmlwidgets::JS(
    "function(settings, json) {$(this.api().table().container()).css({'font-size': '9pt'});}")
  ))
```

--- 

**Table 2.** *Detection functions for cetacean species and taxonomic categories.*  

```{r, echo=TRUE, eval=FALSE, collapse=TRUE, message=FALSE, warning=FALSE}
reporti$table2 
```

```{r, echo=FALSE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE}
reporti$table2 %>% 
DT::datatable(extensions = "FixedColumns",
              options=list(scrollX=TRUE,
                           scrollY='450px',
                           paging = FALSE,
                           fixedColumns = list(leftColumns = 2),
                           initComplete = htmlwidgets::JS(
    "function(settings, json) {$(this.api().table().container()).css({'font-size': '9pt'});}")
  ))
```
  
--- 

**Table 3.** *Estimates of line-transect parameters for cetacean species and taxonomic categories.*  

```{r, echo=TRUE, eval=FALSE, collapse=TRUE, message=FALSE, warning=FALSE}
reporti$table3 
```

```{r, echo=FALSE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE}
reporti$table3 %>% 
DT::datatable(extensions = "FixedColumns",
              options=list(scrollX=TRUE,
                           scrollY='450px',
                           paging = FALSE,
                           fixedColumns = list(leftColumns = 2),
                           initComplete = htmlwidgets::JS(
    "function(settings, json) {$(this.api().table().container()).css({'font-size': '9pt'});}")
  ))
```

--- 

**Table 4.** *Estimates of density (individuals per 1,000 km2) and abundance for cetacean species and taxonomic categories sighted while on systematic survey effort.*  

```{r, echo=TRUE, eval=FALSE, collapse=TRUE, message=FALSE, warning=FALSE}
reporti$table4 
```

```{r, echo=FALSE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE}
reporti$table4 %>% 
DT::datatable(extensions = "FixedColumns",
              options=list(scrollX=TRUE,
                           scrollY='450px',
                           paging = FALSE,
                           fixedColumns = list(leftColumns = 2),
                           initComplete = htmlwidgets::JS(
    "function(settings, json) {$(this.api().table().container()).css({'font-size': '9pt'});}")
  ))
```

--- 

**Table A1.** *Study areas*. 

```{r, echo=TRUE, eval=FALSE, collapse=TRUE, message=FALSE, warning=FALSE}
reporti$tableA1
```

```{r, echo=FALSE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE}
reporti$tableA1 %>% 
DT::datatable(extensions = "FixedColumns",
              options=list(scrollX=TRUE,
                           paging = FALSE,
                           fixedColumns = list(leftColumns = 1),
                           initComplete = htmlwidgets::JS(
    "function(settings, json) {$(this.api().table().container()).css({'font-size': '9pt'});}")
  ))
```

--- 

**Table A2.** *Effort totals and by Beaufort sea-state, in each survey year*. 

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE}
reporti$tableA2
```


### Plots {-}  

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE, fig.width=7, fig.height=21}
lta_plot(species = NULL, 
         lta_result = ltas, 
         years = c(2017, 2020))
```


## Validation {-} 

To validate our results and the routines within `LTabundR`, we will compare the `LTabundR` WHICEAS results to those from A.L. Bradford et al. (2022). To do so, we first bring in their results, which we've staged in a GoogleSheet provided in the code chunk below: 

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE}
library(gsheet)
url <- 'https://docs.google.com/spreadsheets/d/1S94I9I0R589Z8PFP41lYOF8hM6yetap7RFlPdBeet1M/edit?usp=sharing'
alb <- gsheet2tbl(url)
```

Next we format these data to facilitate joining and analyzing alongside our new WHICEAS results within a `tidyverse` framework: 

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
alb <-
    alb %>%
    mutate(title=gsub("'","",species)) %>%
    tidyr::pivot_longer(3:ncol(alb),
                        names_to='column',
                        values_to='value',
                        values_transform = as.character) %>%
    mutate(source='ALB et al. (2022)',
           title = species,
           code = stringr::str_pad(code,width=3, side='left', pad='0'),
           year = ifelse(grepl('17', column), 2017, 2020) %>% as.numeric) %>%
    select(-species) %>%
    rename(species = code) %>%
    mutate(column = gsub('17','',column)) %>%
    mutate(column = gsub('20','',column)) %>%
    group_by(title, species, year) %>%
    summarize(g0 = value[column == 'g0'][1] %>% as.numeric,
              g0_cv = value[column == 'gcv'][1] %>% as.numeric,
              ESW = value[column == 'esw'] %>% as.numeric,
              ss = value[column == 'ss'] %>% as.numeric,
              D = value[column == 'd'] %>% as.numeric,
              N = gsub(',','', value[column == 'n']) %>% as.numeric,
              CV = value[column == 'cv'] %>% as.numeric,
              L95 = gsub(',','', value[column == 'lci']) %>% as.numeric,
              U95 = gsub(',','', value[column == 'uci']) %>% as.numeric) %>%
    filter(!is.na(D)) %>%
    mutate(N = round(N))

# Modify names
names(alb)[4:ncol(alb)] <- paste0('alb_',names(alb)[4:ncol(alb)] )
```

Review these results: 

```{r, echo=TRUE, eval=FALSE, collapse=TRUE, message=FALSE, warning=FALSE}
alb
```

```{r, echo=FALSE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE}
alb %>% 
  DT::datatable(extensions = "FixedColumns",
                options=list(scrollX=TRUE,
                             scrollY='450px',
                             paging = FALSE,
                             initComplete = htmlwidgets::JS(
    "function(settings, json) {$(this.api().table().container()).css({'font-size': '9pt'});}")
  ))
```

### `LTabundR` results {-} 

We will compare the ALB et al. (2022) results to all three versions of the WHICEAS analysis we have produced: (1) manually specified g(0) parameters, (2) Rg(0) estimates from Barlow et al. (2015), and (3) new Rg(0) estimates produced by `LTabundR` automatically. We will bring those results back in and join each to the results from ALB et al. (2022), using a custom helper function, `whiceas_join()`: 

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE, fig.heigh=8, fig.width=8}
whiceas_join <- function(alb, ltas){
  # Combine core info from each LTA list into a dataframe
  ltabundr <- data.frame()
  for(i in 1:length(ltas)){
    lti <- ltas[[i]]
    ltabundi <-
      left_join(lti$estimate,
                lti$bootstrap$summary %>% 
                  select(title, Region, year, species, iterations, 
                         g0_cv, CV, L95, U95),
                by=c('title', 'Region', 'year', 'species'))
    ltabundr <- rbind(ltabundr, ltabundi)
  }

  # Format ltabundr data
  ltabundr <- ltabundr %>%
    mutate(year = as.numeric(year),
           D = D*1000,
           N = round(N)) %>%
    select(title, species, year, g0=g0_est, g0_cv, 
           ESW = ESW_mean, ss = size_mean, D, N, CV, L95, U95)

  # Join ALB to LTabundR
  mr <- left_join(ltabundr, alb, by=c('title', 'year'))
  return(mr)
}
```

Read in, format, and join the results: 

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE, fig.heigh=8, fig.width=8}
# Manually-specified g0
ltas <- lta_enlist('whiceas/lta_manual/')
mr1 <- whiceas_join(alb, ltas) %>% mutate(scenario = 'Manual g(0)')

# Barlow (2015) Rg0
ltas <- lta_enlist('whiceas/lta_barlow/')
mr2 <- whiceas_join(alb, ltas) %>% mutate(scenario = 'Barlow (2015) Rg(0)')

# New auto-generated Rg(0)
ltas <- lta_enlist('whiceas/lta/')
mr3 <- whiceas_join(alb, ltas) %>% mutate(scenario = 'New Rg(0) estimates')

# Join them together
mr <- rbind(mr1, mr2, mr3) %>% 
  mutate(scenario = factor(scenario, levels = c('Manual g(0)',
                                                'Barlow (2015) Rg(0)',
                                                'New Rg(0) estimates')))
```

Now we can compare `LTabundR` estimates to those from ALB et al. (2021): 

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE, fig.heigh=9, fig.width=8}
library(ggplot2)
library(plotly)

# Density
ggplot(mr,
       aes(x=alb_D, y=D, color=title, shape=factor(year))) +
    geom_point() +
    scale_x_continuous(trans='log', limits=c(0.1, 50), 
                       breaks = c(0.01, 0.1, 0.25, 0.5, 1.0, 2.5, 5, 10, 25, 50),
                       labels = function(x)round(x, 2)) +
    scale_y_continuous(trans='log', limits=c(0.1, 50), 
                       breaks = c(0.01, 0.1, 0.25, 0.5, 1.0, 2.5, 5, 10, 25, 50),
                       labels = function(x)round(x, 2)) +
    geom_abline(slope = 1, intercept = 0, lty=3, alpha=.6) +
    ylab('LTabundR') + xlab('ALB et al. (2022)') +
  facet_wrap(~scenario, nrow=3) + 
  labs(title='Density (cetaceans per 1,000 km2)', 
       shape='Year', color = 'Species') + 
  theme_light()
```

&nbsp; 

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE, fig.heigh=9, fig.width=8}
# Abundance
ggplot(mr,
       aes(x=alb_N, y=N, 
           color=title, shape=factor(year))) +
    geom_point() +
    scale_x_continuous(trans='log', 
                       limits=c(10, 30000), 
                       breaks = c(10, 25, 100, 250, 1000, 2500, 10000, 30000),
                       labels = function(x)round(x)) +
    scale_y_continuous(trans='log', 
                       limits=c(10, 30000), 
                       breaks = c(10, 25, 100, 250, 1000, 2500, 10000, 30000),
                       labels = function(x)round(x)) +
    geom_abline(slope = 1, intercept = 0, lty=3, alpha=.6) +
    ylab('LTabundR') + xlab('ALB et al. (2022)') +
  facet_wrap(~scenario, nrow=3) + 
  labs(title='WHICEAS abundance', 
       shape='Year', color = 'Species') + 
  theme_light()
```

&nbsp; 

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE, fig.heigh=9, fig.width=8}
# g(0) estimate
ggplot(mr,
       aes(x=alb_g0, y=g0, 
           color=title, shape=factor(year))) +
    geom_point() +
    geom_abline(slope = 1, intercept = 0, lty=3, alpha=.6) +
    ylab('LTabundR') + xlab('ALB et al. (2022)') +
  facet_wrap(~scenario, nrow=3) + 
  labs(title='Estimates of g(0)', 
       shape='Year', color = 'Species') + 
  theme_light()
```

&nbsp; 

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE, fig.heigh=9, fig.width=8}
# g(0) CV estimate
ggplot(mr,
       aes(x=alb_g0_cv, y=g0_cv, 
           color=title, shape=factor(year))) +
    geom_point() +
    geom_abline(slope = 1, intercept = 0, lty=3, alpha=.6) +
    ylab('LTabundR') + xlab('ALB et al. (2022)') +
  facet_wrap(~scenario, nrow=3) + 
  labs(title='Estimates of CV of g(0)', 
       shape='Year', color = 'Species') + 
  theme_light()
```
