# Appendix: Strata gallery {#stratagallery}

```{r, echo=FALSE, eval=TRUE, collapse=TRUE, include=FALSE}
library(dplyr)
library(magrittr)
library(LTabundR)
```

## Central North Pacific {-} 

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, warning=FALSE, message=FALSE, fig.width=6, fig.height=4}
strata_explore('cnp')
```

To acquire the filepath to one of these strata, pass the index (or indices) printed in the map titles above to the function `strata_select()`:  

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, warning=FALSE, message=FALSE, fig.width=6, fig.height=4}
strata <- strata_select(selections = c(1,3),
                              region = 'cnp')
```

This function returns a named list that can be passed directly to the `strata` argument in `load_settings()`.  

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, warning=FALSE, message=FALSE, fig.width=6, fig.height=4}
strata %>% names
```

The second slot, `$paths`, contains the filepaths you would need to pass to the `load_settings()` function.  


## California Current {-}  

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, warning=FALSE, message=FALSE, fig.width=6, fig.height=6}
strata_explore('ccs')
```

## ETP {-}

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, warning=FALSE, message=FALSE, fig.width=6, fig.height=4}
strata_explore('etp')
```


