# NOAA/NMFS species codes {#spp_codes}

```{r, echo=FALSE, eval=TRUE, collapse=TRUE, include=FALSE}
library(dplyr)
library(magrittr)
library(LTabundR)
library(knitr)
```

## Table {-} 

`LTabundR` provides the standard table of NOAA/NMFS species codes as a built-in dataset:

```{r}
data(species_codes)
```

```{r, echo = FALSE}
species_codes %>% 
  DT::datatable(options=list(initComplete = htmlwidgets::JS(
          "function(settings, json) {$(this.api().table().container()).css({'font-size': '9pt'});}")
       ))
```

This version of the species codes table was provided by Amanda Bradford (Pacific Islands Fisheries Science Center) in 2021.  

For an easier way to find a species of interest, look into the `species_translator()` function on the [Miscellaneous functions](#misc_functions) page.  


## `species_translator()` {-} 

To streamline the management of species codes, scientific names, common names, etc., in the functions throughout this package, we have developed a "translator" function that returns the various identifiers for a species according to a variety of search terms. 

You can search by species code:  

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE}
# source('R/species_translator.R`)
species_translator(id = '035') %>% t()
```

By the short code name:  

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE}
species_translator(id = 'LONG_PILOT') %>% t()
```

By the scientific name:  

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE}
species_translator(id = 'Globicephala melas') %>% t()
```

Or by one of the species' common names:  

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE}
species_translator(id = 'Long-finned pilot whale') %>% t()
```

The function will return any species for which there is a partial match:  

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE}
species_translator(id = 'megap') %>% t()
```

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE}
species_translator(id = 'killer')
```


Note that if `species_codes` is `NULL`, as in the examples above, the list of codes used in `ABUND9` will be used as a default.  
