# Miscellaneous functions {#misc_functions}

```{r, echo=FALSE, eval=TRUE, collapse=TRUE, include=FALSE}
library(dplyr)
library(magrittr)
library(LTabundR)
```

### `species_translator()` {-} 

To streamline the management of species codes, scientific names, common names, etc., in the functions throughout this package, we have developed a "translator" function that returns the various identifiers for a species according to a variety of search terms. 

You can search by species code:  

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE}
# source('R/species_translator.R`)
species_translator(id = '035') %>% t()
```

By the short code name:  

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE}
species_translator(id = 'LONG_PILOT') %>% t()
```

By the scientific name:  

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE}
species_translator(id = 'Globicephala melas') %>% t()
```

Or by one of the species' common names:  

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE}
species_translator(id = 'Long-finned pilot whale') %>% t()
```

The function will return any species for which there is a partial match:  

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE}
species_translator(id = 'megap') %>% t()
```

```{r, echo=TRUE, eval=TRUE, collapse=TRUE, message=FALSE, warning=FALSE}
species_translator(id = 'killer')
```


Note that if `species_codes` is `NULL`, as in the examples above, the list of codes used in `ABUND9` will be used as a default.  

